Efficient Phenotype Prediction Using Incomplete Omics Data

### 2017-10-16 paper/manuscript.tex
paper, edits from meeting on 2017_10_12

%% THE MANUSCRIPT TITLE
\title{Omics-based Single Step Trait Prediction}

@@

\begin{abstract}
Predicting genotypic values with great precision is pivotal for effective
candidate selections in animal and plant breeding.
Novel `omics' predictors have shown to improve upon established genome-based
predictions of some complex traits but yet require laborious and expensive
assays.
In animal breeding, single-step prediction was introduced to efficiently use
pedigree information, collected on a large number of animals, and genomic
information, collected on a smaller subset of animals, simultaneously for breeding
value estimation without bias.
Single-step prediction has not yet been attempted with predictors calculated from
quantitative features displaying heritabilities smaller than one.
Using two maize datasets of inbred lines and hybrids we show that transcriptomic
data can be imputed successfully, boosting forecasts when their predictive
ability exceeds that of pedigree or genomic data.
Covering only a subset of genotypes with `omics' predictors and imputing all
others using pedigree or genomic data could enable breeders to improve trait
predictions while keeping costs under control.
Employing `omics' predictors could particularly improve candidate selection in
hybrid breeding where the success of forecasts is a strongly convex function of
predictive ability.
\end{abstract}

@@

We wish to submit the enclosed manuscript ``Omics-based Single Step Prediction''
by Westhues et al.\@ for publication in TAG.

Recent studies in maize inbred lines~\cite{Guo2016}, maize hybrids
\cite{Zenke-Philippi2017,Westhues2017}, rice hybrids~\cite{Dan2016,Xu2016} and
humans~\cite{Vazquez2016} have suggested that `omics' predictors, such as the
metabolome and the transcriptome, can improve the prediction of traits with
complex genetic architectures over that achieved with pedigree or genomic data.
Particularly hybrid breeding, where the success of forecasts increases
exponentionally with predictive abilities, could benefit from the use of
predictors that yield more precise estimates of candidate performance.
Currently, the adoption of these novel predictors by breeders is impeded
by their higher sampling costs compared to pedigree and genomic data.
Here, we present single-step prediction of complex traits as a framework for
combining pedigree and/or genomic data with other `omics' predictors collected
on a subset of all available genotypes to boost predictive abilities while
managing costs.

Using two extensive phenotypic data sets from a public hybrid silage maize
breeding program and a maize diversity panel, we a) investigate the success of
imputing a quantitative predictor with feature heritabilities less than one and
b) the influence of the fraction of genotypes covered by two predictors on
predictive abilities in single step prediction.

@@

\begin{enumerate}
\item In both data sets, single-step prediction improved predictive abilities
	over those of the predictor covering all genotypes when the discrepancy 
	between the	two predictors in terms of predictive ability were moderate to
	large in favor of the incomplete predictor.
\item When 20\%-30\% ($n \approx 25$) of the genotypes were covered by both
	predictors, single step-based predictive abilities converged to those
	obtained with the incomplete predictor.
\end{enumerate}

In conclusion our results demonstrate that single-step prediction, using
an established predictor such as pedigree information, can benefit from sampling
a highly performant predictor, such as transcriptomic data, on a subset of all
genotypes.
The framework presented herein should help promote the use of attractive
`omics' predictors by balancing their advantages with their currently higher
costs.

% Introduction
\section{Introduction}
% Rationale for predictions.
Genomic prediction, pioneered by \citeNP{Meuwissen2001}, has revolutionized
animal and plant breeding by a more efficient selection of promising candidates 
without phenotypic records~\cite{DeLosCampos2013,Garcia-Ruiz2016}.
Its use is mainly motivated by the high costs and oftentimes time-consuming
phenotyping, which can be reduced when all genotypes are covered by a predictor 
while limiting phenotyping to a few genotypes with high precision
\cite{Kadam2016}.
The performance of all other genotypes is then forecast using model parameters
estimated in the subset of individuals with full information.
In animal breeding, genomic selection has increased the annual selection gain 
by $\approx$ 50--100\% for yield traits and by 300--400\% for traits with low 
heritability~\cite{Garcia-Ruiz2016}.

@@

\section{Materials and Methods}
\Genetics2level{Exp1}
To address objectives (i) and (ii) a published data set of maize inbred lines
\cite{Yang2014} was used.
It originally comprised 513 maize lines representing the global maize
diversity  and was reduced to the set of tropical and subtropical lines
($n = 211$), subsequently referred to as Exp1, which is the largest of the four
subgroups classified by~\citeN{Guo2016}.
All inbred lines were evaluated in five different environments in China
described in detail by~\citeNP{Yang2014}.
Best linear unbiased predictors (BLUPs) were calculated by~\citeN{Yang2014} for
all 211 inbred lines and for 17 traits of which six are presented in this study 
(Table~\ref{table:TraitKey}).
The selection of these traits was done based on our prediction results so that
two traits were predicted with greater precision by transcriptomic data and four
further traits were predicted with greater precision by genomic data.
After filtering for minor allele frequency ($\geq 5$\%), heterozygosity rate
($\leq 5$\%) and call frequency ($\geq 95$\%), missing genotypes were imputed
using the Beagle software~\cite{Browning2009} resulting in 37,760 SNPs for these
211 genotypes.
A subset of 28,850 annotated genes was available for 149 out of the 211
genotypes~\cite{Fu2013} and provided the basis for further analyses.

\subsection{Population structure}
Population structure in Exp1 was examined through a STRUCTURE
\cite{Pritchard2000} analysis with the number of putative ancestral
populations $K$ ranging from 2 to 9.
Based on the estimated likelihood of the data for each $K$, we chose $K=4$
and carried out a principal component analysis (PCA) on the scaled and centered
genomic feature matrix using the \emph{pca} function from the \emph{R}-package
\emph{LEA}~\cite{Frichot2015}.
Least-squares estimates of ancestry proportions were estimated based on sparse 
nonnegative matrix factorization algorithms~\cite{Frichot2014} using the 
\emph{snmf} function in \emph{LEA}.
Genotypes in the PCA plot (Fig.~\ref{fig:inbred-pca}) were colorized based on their
primary assignment to any of the four putative ancestral populations.
To mitigate the possibility of inflated predictive abilities, which can arise
from large differences in the average performance of the different
subpopulations~\cite{Windhausen2012}, we reduced the panel of inbred lines to
164 genotypes, which were assigned to the largely overlapping clusters 1 and 4 in
Fig.~\ref{fig:inbred-pca}.
In summary, genomic and phenotypic information were available for 164 inbred
lines, which we denote as the `Full' set whereas transcriptomic data were available
for a subset of 110 inbred lines, which we denote as `Reduced'.

@@

\Genetics2level{Exp2}
To address objectives (i) and (iii), a maize hybrid data set, subsequently
denoted as Exp2, was used.
The material comprised 1,521 hybrids produced in 16 factorial mating designs
between 142 Dent and 103 Flint parent lines described in detail by
\citeN{Westhues2017}.
Best linear unbiased estimates (BLUEs) were computed across all factorials and 
three or more agro-ecologically diverse environments across Germany for seven
agronomically important traits in silage maize by accounting for year, location,
field replication, block and genotype effects as well as their interactions
(Table~\ref{table:TraitKey}).
All 245 parent lines have pedigree records back to the generation of their 
grandparents or deeper~\cite{Westhues2017} and were genotyped using the Illumina 
BeadChip MaizeSNP50~\cite{Ganal2011}.
The same SNP quality checks as for the material in Exp1 were applied here,
yielding 7,013 polymorphic marker loci for the Dent and 6,212 for the Flint
lines, respectively.
Gene expression data for seedlings seven days after sowing were obtained for a 
subset of 60 Dent and 43 Flint lines, that are parents of 685 hybrid progeny.
These data were obtained through two-color hybridizations using a custom 2K 
microarray (GPL22267)~\cite{Westhues2017}.
After applying established normalization procedures~\cite{Smyth2003,Ritchie2007},
BLUEs for gene expression were computed for 1,323 transcripts
\cite{Westhues2017} using the R-package \emph{limma}~\cite{Ritchie2015a}.

% Please add the following required packages to your document preamble:
% \usepackage{booktabs}
\begin{table}[]
\centering
\caption{Abbreviations for the trait names}
\label{table:TraitKey}
\begin{tabular}{@{}llll@{}}
\toprule
Trait              & Abbreviation & Data Set  & Experiment \\ \midrule
Cob weight         & CW           & Diversity & Exp1       \\
Days to silking    & DS           & Diversity & Exp1       \\
Ear diameter       & ED           & Diversity & Exp1       \\
100-grain weight   & GW           & Diversity & Exp1       \\
Kernel width       & KW           & Diversity & Exp1       \\
Plant height       & PH           & Diversity & Exp1       \\
Dry matter yield   & DMY          & Hybrid    & Exp2       \\
Dry matter content & DMC          & Hybrid    & Exp2       \\
Lignin             & ADL          & Hybrid    & Exp2       \\
Fat                & FAT          & Hybrid    & Exp2       \\
Protein            & PRO          & Hybrid    & Exp2       \\
Starch             & STA          & Hybrid    & Exp2       \\
Sugar              & SUG          & Hybrid    & Exp2       \\ \bottomrule
\end{tabular}
\end{table}

\Genetics2level{Prediction}
\subsection{Kernels}
Depending on the data set, up to three predictors were available for agronomic
trait predictions, namely pedigree data (P), genomic data (G) and
transcriptomic data (T).
The corresponding feature matrix of the inbred lines ($\mathbf{M}$) has
dimensions $n \times p$ where $n$ pertains to the number of genotypes and $p$
pertains to the number of features; that is transcripts for T and SNPs for G.
For the sake of simplicity the derivations are detailed first for predicting
breeding values of inbred lines.
Later, we briefly describe the extension to models required for predicting the
genetic value of hybrids.
For the complete predictor we generated kernels by centering and standardizing
all features in $\mathbf{M}$ to unit variance:

\begin{equation} \label{eq:GenomicRelationship}
  \mathbf{K} = \frac{1}{M} \mathbf{M} \mathbf{M}^{\top},
\end{equation}

where $M$ denotes the number of features~\cite{VanRaden2008}.
In the case of pedigree data (P), the numerator relationship matrix was used
directly for $\mathbf{K}$.

@@

where $\mathbf{y}$ is the vector of observed inbred line performance $\mu$ is
the fixed model intercept, and $\mathbf{e}$ is the residual error.
For the inbred lines we fit random genotype effects $\mathbf{g}$, which are
associated with the vector of phenotypic performance $\mathbf{y}$ through a
corresponding design matrix $\mathbf{Z}$.
The random effects ($\mathbf{g}$) have expectation zero and variance equal to
$\mathbf{K} \sigma^{2}_{g}$ and $\mathbf{I} \sigma^2_{e}$ for the residual
error, respectively.
Here, we did not consider the inclusion of dominance effects for the hybrid
material after a previous investigation into the predictive ability of various
``omics'' predictors in the same material concluded no considerable gains 
\cite{Westhues2017}.
All prediction models were implemented using the R-packages \textit{BGLR}
\cite{Perez2014} and \textit{sspredr}
\url{https://github.com/mwesthues/sspredr}.

\subsection{Single-step prediction}
If the breeding values are fully captured by the features in $\mathbf{M}$,
Eq.~\ref{eq:KBLUPModel} can be expressed as~\cite{Fernando2014}

@@

Consider now the situation in which one predictor is complete in the sense that
the whole population is covered by features for that predictor but only a subset
of that population has information for a second predictor.
We can take a similar route as in \citeNP{Fernando2014} and impute covariates 
of the incomplete predictor, denoted as `predictor' and indicated through the
asterisk symbol `*', by using covariates of the complete predictor, denoted as
`imputor'.
Let the subscript $1$ denote individuals covered only by the predictor whereas
individuals covered by both, the imputor and the predictor, are indicated by the subscript $2$.

Let us further denote the matrix of centered covariates of the predictor as
\begin{equation} \label{eq:incomplete-covariates}
\tensor[]{\mathbf{W}}{^{*}} =
\begin{bmatrix}
  \mathbf{Z}_{1} \; \tensor[]{\mathbf{\hat{W}}}{^{*}_{1}} \\
  \tensor[]{\mathbf{W}}{^{*}_{2}}
 \end{bmatrix},
\end{equation}
where observed feature covariates are stored in $\tensor[]{\mathbf{W}}{^{*}_{2}}$
while the feature covariates for genotypes not covered by the predictor can be estimated/imputed with the aid of the covariance matrix of the
imputor $\mathbf{K}$ and the feature matrix of the predictor 
$\tensor[]{\mathbf{W}}{^{*}_{2}}$ by calculating
($\tensor[]{\mathbf{\hat{W}}}{^{*}_{1}} = \tensor[]{\mathbf{K}}{_{12}} \;
\tensor[]{\mathbf{K}}{_{22}^{-1}} \; \tensor[]{\mathbf{W}}{^{*}_{2}}$).
In this case we need to partition the vector of breeding values $\mathbf{g}$
into a component $\mathbf{g}_{1}$ (\textit{i.e.,} breeding values of genotypes
covered only by the imputor) and a component
$\mathbf{g}_{2} = \tensor[]{\mathbf{W}}{^{*}_{2}} \; 
\tensor[]{\boldsymbol{\alpha}}{^{*}}$ (\textit{i.e.,} genetic
values of genotypes covered by both predictors, where
$\tensor[]{\boldsymbol{\alpha}}{^{*}}$ pertains to the random feature effects of
the predictor).
Features are distributed according to a multivariate normal distribution:
\begin{align} \label{eq:mrna1}
	\mathbf{g_1} &= E(\mathbf{g}_1|\mathbf{g}_2) +
	\boldsymbol{\epsilon} \\
	&= \tensor[]{\mathbf{K}}{_{12}} \;
       	\tensor[]{\mathbf{K}}{_{22}^{-1}} \;
       	\tensor[]{\mathbf{W}}{^{*}_{2}} \;
	\tensor[]{\boldsymbol{\hat{\alpha}}}{^{*}} +
       	(\mathbf{g_1} - \tensor[]{\mathbf{K}}{_{12}} \;
	\tensor[]{\mathbf{K}}{_{22}^{-1}} \;
       	\tensor[]{\mathbf{W}}{^{*}_{2}} \;
	\tensor[]{\boldsymbol{\hat{\alpha}}}{^{*}}) \\
	&= \mathbf{\hat{g}}_1 + \boldsymbol{\epsilon},
\end{align}

@@

The structure of the residual imputation/prediction error is known, and
therefore, can be modeled.

@@

The model above indicates that single-step prediction models ultimately use
only the matrix of known and imputed feature covariates $\mathbf{W}$ for the
estimation of breeding values while accounting for the imputation error
$\boldsymbol{\epsilon}$.

@@

\Genetics2level{Core sampling}
In order to evaluate the influence of the fraction of genotypes that is not
covered by the incomplete predictor on predictive ability we generated core
sets of nine different sizes from Exp1 (Fig.~\ref{fig:compound-plot}~a).
The base of each of these core sets was the set of 110 genotypes that were
covered by both, genomic and transcriptomic data.
A core set was created by artificially declaring a pre-specified fraction 
$x = {10\%, 20\%, \dots, 90\%}$ of genotypes as lacking of transcriptomic 
information.
For each value of $x$ we created 100 core sets $s \in {1, 2, 3, \dots, 100}$ by
repeatedly sampling genotypes, whose trancsriptomic information was declared
as missing, at random (Fig.~\ref{fig:compound-plot}~b).
The artificially removed transriptomic information on these genotypes was then
imputed.
For the each core set, comprising 110 genotypes, predictive abilities were again
computed as described above, using LOOCV and bootstrapping.

@@

\section*{Results}
\Genetics2level{Exp1}
At the level of the 110 genotypes for which both, genomic (G) and 
transcriptomic (T) information were available, the use of T yielded higher
predictive abilities than the use of G for two out of six traits
(Fig.~\ref{fig:InbredResults}A).
In terms of predictive ability, the absolute advantage of T over G was as large 
as 0.15, observed for the trait ED\@.
The predictive ability for predictions based on G were always higher for the
full set of inbred lines ($n = 164$) than for the reduced set of inbred lines
($n = 110$).
The combination of G and T in a single step (\textit{i.e.}, $G|T^{*}$, where
$G$ is the imputor and $T^{*}$ is the predictor, for the prediction of all 164
lines, performed at least as well as G alone when T was superior over G for the
reduced set of 110 lines.
In the case of KW and ED, the use of $G|T^{*}$ yielded a slight improvement in
predictive ability over G.
Where T was a worse predictor for the reduced set of 110 lines than G,
predictive abilities of $G|T^{*}$ for the full set of 164 lines were also worse
than those based on G alone.

\begin{figure}[H]
  \centering
  \includegraphics{./tables_figures/pred_ability_inbred.pdf}
  \caption{
  Predictive abilities and average, bootstrapped standard errors for
  tropical/subtropical inbred lines from the maize diversity panel and six
  agronomic traits.
  As predictors, genomic (G) data, transcriptomic (T) data and their 
  combination ($\tensor[_{C}]{G}{}\tensor[_{I}]{T}{}$) were used.
  The `Reduced' panel includes a subset of 110 lines covered by genomic and
  transcriptomic information whereas the `Full' panel comprises all 164
  genotypes.
  In the `Full' set of genotypes, transcriptomic records were imputed for 54
  lines.
  }
\label{fig:InbredResults}
\end{figure}

\Genetics2level{Exp2}
In the `Reduced' set of Exp2, transcriptomic data were the best single predictor
for the traits ADL, PRO and particulaly for DMY where the predictive ability
obtained with T was 0.08 points higher than for G as the second best predictor for
this trait (Fig.~\ref{fig:HybridResults}).
In the reduced set, P had the highest predictive ability for STA and DMC\@.
For FAT and SUG, G was the superior predictor inside the reduced set.
In the full set of 1,521 hybrids, predictive abilities obtained with G increased
markedly for all traits with the exception of FAT\@.
The single-step combination of genomic and transriptomic information
($G|T^{*}$) yielded a small improvement over G alone for the traits DMY and PRO\@.
Except for PRO, compared to using only G, trait prediction did not benefit from 
the single-step approach when T was not superior to G in the reduced set.
Predictive abilities resulting from combinations of pedigree with genomic
($P|G^{*}$) information were better than P alone for all traits and $P|T^{*}$
improved upon P in the `full' set for four traits.

>>>>

%% THE MANUSCRIPT TITLE
\title{Efficient Phenotype Prediction Using Incomplete Omics Data}

@@

\section{Key message}
Borrowing information from a highly performant `omics' predictor, sampled on a
subset of inbred lines, can efficiently improve the prediction of complex
traits.

\begin{abstract}
Predicting genotypic values with great precision is pivotal for effective
selection of candidates in animal and plant breeding.
Novel `omics' predictors have shown to improve upon established genome-based
predictions of some complex traits but yet require laborious and expensive
assays.
In animal breeding, single-step prediction was introduced to efficiently use
pedigree information, collected on a large number of animals, and genomic
information, collected on a smaller subset of animals, simultaneously for breeding
value estimation without bias.
Single-step prediction has not yet been attempted with predictors calculated from
quantitative features displaying repeatabilities smaller than one.
Using two maize datasets of inbred lines and hybrids we show that transcriptomic
data can be imputed successfully, boosting forecasts when their predictive
ability exceeds that of pedigree or genomic data.
Our results suggest that covering only a subset of inbred lines with `omics'
predictors and imputing all others using pedigree or genomic data could enable
breeders to improve trait predictions while keeping costs under control.
Employing `omics' predictors could particularly improve candidate selection in
hybrid breeding where the success of forecasts is a strongly convex function of
predictive ability.
\end{abstract}

@@

We wish to submit the enclosed manuscript 
``Efficient Phenotype Prediction Using Incomplete Omics Data'' by Westhues et 
al.\@ for publication in TAG.

Recent studies in maize inbred lines~\cite{Guo2016}, maize hybrids
\cite{Zenke-Philippi2017,Westhues2017} and rice hybrids~\cite{Dan2016,Xu2016}
have suggested that `omics' predictors, such as the metabolome and the
transcriptome, can improve the prediction of traits with complex genetic 
architectures over that achieved with pedigree or genomic data.
Particularly hybrid breeding, where the success of forecasts increases
exponentionally with predictive abilities, could benefit from the use of
predictors that yield more precise estimates of candidate performance.
Currently, the adoption of these novel predictors by breeders is impeded
by their higher costs compared to pedigree and genomic data.
Here, we present single-step prediction of complex traits as a framework for
combining pedigree and/or genomic data with other `omics' predictors collected
on a subset of all available inbred lines to boost predictive abilities while
controlling costs.

Using two extensive phenotypic data sets from a public hybrid silage maize
breeding program and a maize diversity panel published in TAG~\cite{Westhues2017,
Guo2016}, we a) investigate the success of imputing a quantitative predictor
with feature repeatabilities less than one and b) the influence of the fraction
of inbred lines covered by two predictors on predictive abilities in prediction
based on incomplete `omics' data.

@@

\begin{enumerate}
\item In both data sets, single-step prediction improved predictive abilities
	over those of the predictor covering all inbred lines when the
	discrepancy between the	two predictors in terms of predictive ability
	were moderate to large in favor of the incomplete predictor.
\item When 20\%-30\% ($n \approx 25$) of the inbred lines were covered by both
	predictors, single step-based predictive abilities converged to those
	obtained with the incomplete predictor.
\end{enumerate}

In conclusion our results demonstrate that single-step prediction, using
an established predictor such as pedigree information, can benefit from sampling
a highly performant predictor, such as transcriptomic data, on a subset of all
inbred lines.
The framework presented herein should help promote the use of attractive
`omics' predictors by balancing their advantages with their currently higher
costs.

% Introduction
\section{Introduction}
% Rationale for predictions.
Genomic prediction, pioneered by \citeNP{Meuwissen2001}, has revolutionized
animal and plant breeding by a more efficient selection of promising candidates 
without phenotypic records~\cite{DeLosCampos2013,Garcia-Ruiz2016}.
Its use is mainly motivated by the high costs and oftentimes time-consuming
phenotyping.
Costs can be reduced when all genotypes, equivalent to inbred lines in plant
breeding and animals in animal breeding, are covered by a predictor while
limiting phenotyping to a few genotypes with high precision \cite{Kadam2016}.
The performance of all other genotypes is then forecast using model parameters
estimated in the subset of individuals with full information.
In animal breeding, genomic selection has increased the annual selection gain 
by $\approx$ 50--100\% for yield traits and by 300--400\% for traits with low 
heritability~\cite{Garcia-Ruiz2016}.

@@

\section{Materials and Methods}
\Genetics2level{Experiment 1}
To address objectives (i) and (ii) a published data set of maize inbred lines
\cite{Yang2014} was used.
It originally comprised 513 maize lines representing the global maize
diversity  and was reduced to the set of tropical and subtropical lines
($n = 211$), subsequently referred to as Exp1, which is the largest of the four
subgroups classified by~\citeN{Guo2016}.
All inbred lines were evaluated in five different environments in China
described in detail by~\citeNP{Yang2014}.
Best linear unbiased predictors (BLUPs) were calculated by~\citeN{Yang2014} for
all 211 inbred lines and for 17 traits of which six are presented in this study.
The selection of these traits was done based on our prediction results so that
two traits were predicted with greater precision by transcriptomic data and four
further traits were predicted with greater precision by genomic data.
After filtering for minor allele frequency ($\geq 5$\%), heterozygosity rate
($\leq 5$\%) and call frequency ($\geq 95$\%), missing genotypes were imputed
using the Beagle software~\cite{Browning2009} resulting in 37,760 SNPs for these
211 genotypes.
A subset of 28,850 annotated genes was available for 149 out of the 211
genotypes~\cite{Fu2013} and provided the basis for further analyses.

\subsection{Population structure}
Population structure in Exp1 was examined through a STRUCTURE
\cite{Pritchard2000} analysis with the number of putative ancestral
populations $K$ ranging from 2 to 9.
Based on the estimated likelihood of the data for each $K$, we chose $K=4$
and carried out a principal component analysis (PCA) on the scaled and centered
genomic feature matrix using the \emph{pca} function from the \emph{R}-package
\emph{LEA}~\cite{Frichot2015}.
Least-squares estimates of ancestry proportions were estimated based on sparse 
nonnegative matrix factorization algorithms~\cite{Frichot2014} using the 
\emph{snmf} function in \emph{LEA}.
Genotypes in the PCA plot (Fig.~\ref{fig:inbred-pca}) were colorized based on their
primary assignment to any of the four putative ancestral populations.
To mitigate the possibility of inflated predictive abilities, which can arise
from large differences in the average performance of the different
subpopulations~\cite{Windhausen2012}, we reduced the panel of inbred lines to
164 genotypes, which were assigned to the largely overlapping clusters 1 and 4 in
Fig.~\ref{fig:inbred-pca}.
In summary, genomic and phenotypic information were available for 164 inbred
lines, which we denote as the `Full' set, whereas transcriptomic data were
available for a subset of 110 inbred lines, which we denote as `Reduced'.

@@

\Genetics2level{Experiment 2}
To address objectives (i) and (iii), a maize hybrid data set, subsequently
denoted as Exp2, was used.
The material comprised 1,521 hybrids produced in 16 factorial mating designs
between 142 Dent and 103 Flint parent lines described in detail by
\citeN{Westhues2017}.
Best linear unbiased estimates (BLUEs) were computed across all factorials and 
three or more agro-ecologically diverse environments across Germany for seven
agronomically important traits in silage maize by accounting for year, location,
field replication, block and genotype effects as well as their interactions.
All 245 parent lines have pedigree records back to the generation of their 
grandparents or deeper~\cite{Westhues2017} and were genotyped using the Illumina 
BeadChip MaizeSNP50~\cite{Ganal2011}.
The same SNP quality checks as for the material in Exp1 were applied here,
yielding 7,013 polymorphic marker loci for the Dent and 6,212 for the Flint
lines, respectively.
Gene expression data for seedlings seven days after sowing were obtained for a 
subset of 60 Dent and 43 Flint lines, that are parents of 685 hybrid progeny.
These data were obtained through two-color hybridizations using a custom 2K 
microarray (GPL22267)~\cite{Westhues2017}.
After applying established normalization procedures~\cite{Smyth2003,Ritchie2007},
BLUEs for gene expression were computed for 1,323 transcripts
\cite{Westhues2017} using the R-package \emph{limma}~\cite{Ritchie2015a}.

\Genetics2level{Prediction}
\subsection{Kernels}
Depending on the data set, up to three predictors were available for agronomic
trait predictions, namely pedigree data (P), genomic data (G) and
transcriptomic data (T).
The corresponding feature matrix of the inbred lines ($\mathbf{M}$) has
dimensions $n \times p$ where $n$ pertains to the number of genotypes and $p$
pertains to the number of features; that is transcripts for T and SNPs for G.
For the sake of simplicity the derivations are detailed first for predicting
breeding values of inbred lines.
Later, we briefly describe the extension to models required for predicting the
genetic value of hybrids.
For the complete predictor we generated kernels by centering and standardizing
all features in $\mathbf{M}$ to unit variance~\cite{VanRaden2008}:

\begin{equation} \label{eq:GenomicRelationship}
  \mathbf{K} = \frac{1}{p} \mathbf{M} \mathbf{M}^{\top}.
\end{equation}

In the case of pedigree data (P), the numerator relationship matrix was used
directly for $\mathbf{K}$.

@@

where $\mathbf{y}$ is the vector of observed inbred line performance $\mu$ is
the fixed model intercept, and $\mathbf{e}$ is the residual error.
For the inbred lines we fit random genotype effects $\mathbf{g}$, which are
associated with the vector of phenotypic performance $\mathbf{y}$ through a
corresponding design matrix $\mathbf{Z}$.
The random effects ($\mathbf{g}$) have expectation zero and variance equal to
$\mathbf{K} \sigma^{2}_{g}$ and $\mathbf{I} \sigma^2_{e}$ for the residual
error, respectively.
All prediction models were implemented using the R-packages \textit{BGLR}
\cite{Perez2014} and \textit{sspredr}
\url{https://github.com/mwesthues/sspredr}.

\subsection{Single-step prediction}
If the breeding values are fully captured by the features in $\mathbf{M}$,
Eq.(~\ref{eq:KBLUPModel}) can be expressed as~\cite{Fernando2014}

@@

Consider now the situation in which one predictor is complete in the sense that
the whole population is covered by features for that predictor but only a subset
of that population has information for a second predictor.
We can take a similar route as in~\citeNP{Fernando2014} and impute covariates 
of the incomplete predictor, denoted as `predictor' and indicated through the
asterisk symbol `*', by using covariates of the complete predictor, denoted as
`imputor'.
Let the subscript $1$ denote individuals covered only by the predictor whereas
individuals covered by both, the imputor and the predictor, are indicated by the subscript $2$.

Let us further denote the matrix of centered covariates of the predictor as
\begin{equation} \label{eq:incomplete-covariates}
\tensor[]{\mathbf{W}}{^{*}} =
\begin{bmatrix}
  \mathbf{Z}_{1} \; \tensor[]{\mathbf{\hat{W}}}{^{*}_{1}} \\
  \tensor[]{\mathbf{W}}{^{*}_{2}}
 \end{bmatrix},
\end{equation}
where observed feature covariates are stored in $\tensor[]{\mathbf{W}}{^{*}_{2}}$
while the feature covariates for genotypes not covered by the predictor can be
imputed with the aid of the covariance matrix of the imputor $\mathbf{K}$ and
the feature matrix of the predictor $\tensor[]{\mathbf{W}}{^{*}_{2}}$ by
calculating
($\tensor[]{\mathbf{\hat{W}}}{^{*}_{1}} = \tensor[]{\mathbf{K}}{_{12}} \;
\tensor[]{\mathbf{K}}{_{22}^{-1}} \; \tensor[]{\mathbf{W}}{^{*}_{2}}$).
In this case we need to partition the vector of breeding values $\mathbf{g}$
into a component $\mathbf{g}_{1}$ (\textit{i.e.,} breeding values of genotypes
covered only by the imputor) and a component
$\mathbf{g}_{2} = \tensor[]{\mathbf{W}}{^{*}_{2}} \; 
\tensor[]{\boldsymbol{\alpha}}{^{*}}$ (\textit{i.e.,} genetic
values of genotypes covered by both predictors, where
$\tensor[]{\boldsymbol{\alpha}}{^{*}}$ pertains to the random feature effects of
the predictor).
Features are distributed according to a multivariate normal distribution:
\begin{align} \label{eq:mrna1}
	\mathbf{g_1} &= E(\mathbf{g}_1|\mathbf{g}_2) +
	\boldsymbol{\epsilon} \\
	&= \tensor[]{\mathbf{K}}{_{12}} \;
       	\tensor[]{\mathbf{K}}{_{22}^{-1}} \;
       	\tensor[]{\mathbf{W}}{^{*}_{2}} \;
	\tensor[]{\boldsymbol{\hat{\alpha}}}{^{*}} +
       	(\mathbf{g_1} - \tensor[]{\mathbf{K}}{_{12}} \;
	\tensor[]{\mathbf{K}}{_{22}^{-1}} \;
       	\tensor[]{\mathbf{W}}{^{*}_{2}} \;
	\tensor[]{\boldsymbol{\hat{\alpha}}}{^{*}}) \\
	&= \mathbf{\hat{g}}_1 + \boldsymbol{\epsilon},
\end{align}

@@

The structure of the residual imputation error is known, and therefore, can be
modeled.

@@

The model above indicates that single-step prediction models ultimately use
only the matrix of known and imputed feature covariates $\mathbf{W}^{*}$ for the
estimation of breeding values while accounting for the imputation error
$\boldsymbol{\epsilon}$.

@@

\Genetics2level{Core sampling}
In order to evaluate how prediction accuracy is influenced by the fraction of
genotypes not covered by the incomplete predictor on predictive ability we
generated core sets of nine different sizes from Exp1
(Fig.~\ref{fig:compound-plot}~a).
The base of each of these core sets was the set of 110 genotypes that were
covered by both, genomic and transcriptomic data.
A core set was created by artificially declaring a pre-specified fraction 
$x = {10\%, 20\%, \dots, 90\%}$ of genotypes as lacking of transcriptomic 
information.
For each value of $x$ we created 100 core sets $s \in {1, 2, 3, \dots, 100}$ by
repeatedly sampling genotypes, whose trancsriptomic information was declared
as missing, at random (Fig.~\ref{fig:compound-plot}~b).
The artificially removed transriptomic information on these genotypes was then
imputed.
For the each core set, comprising 110 genotypes, predictive abilities were again
computed as described above, using LOOCV and bootstrapping.

@@

\Genetics2level{Data availability}
The code for the analyses can be accessed from 
\url{https:://github.com/mwesthues/single\_step} and
\url{https://github.com/mwesthues/sspredr}.
All data from Exp1 have been published before and are publicly available.
Data for Exp2 can be made available upon request.

\section*{Results}
\Genetics2level{Experiment 1}
At the level of the 110 genotypes for which both, genomic (G) and 
transcriptomic (T) information were available, the use of T yielded higher
predictive abilities than the use of G for two out of six traits
(Fig.~\ref{fig:InbredResults}A).
In terms of predictive ability, the absolute advantage of T over G was as large 
as 0.15, observed for the trait ear diameter\@.
The predictive ability for predictions based on G were always higher for the
full set of inbred lines ($n = 164$) than for the reduced set of inbred lines
($n = 110$).
The combination of G and T in a single step (\textit{i.e.}, $T^{*}|G$, where
$G$ is the imputor and $T^{*}$ is the predictor, for the prediction of all 164
lines, performed at least as well as G alone when T was superior over G for the
reduced set of 110 lines.
In the case of kernel width and ear diameter, the use of $T^{*}|G$ yielded a slight 
improvement in predictive ability over G.
For the other four traits, where T was a worse predictor for the reduced set of
110 lines than G, predictive abilities of $T^{*}|G$ for the full set of 164
lines were also worse than those based on G alone.

\begin{figure}[H]
  \centering
  \includegraphics{./tables_figures/pred_ability_inbred.pdf}
  \caption{
  Predictive abilities and average, bootstrapped standard errors for
  tropical/subtropical inbred lines from the maize diversity panel and six
  agronomic traits.
  As predictors, genomic (G) data, transcriptomic (T) data and their 
  combination ($T^{*}|G$) were used.
  The `Reduced' panel includes a subset of 110 lines covered by genomic and
  transcriptomic information whereas the `Full' panel comprises all 164
  genotypes.
  In the `Full' set of genotypes, transcriptomic records were imputed for 54
  lines.
  }
\label{fig:InbredResults}
\end{figure}

\Genetics2level{Experiment 2}
In the `Reduced' set of Exp2, transcriptomic data were the best single predictor
for the traits lignin, protein and particulaly for dry matter yield where the
predictive ability obtained with T was 0.08 points higher than for G as the
second best predictor for this trait (Fig.~\ref{fig:HybridResults}).
In the reduced set, P had the highest predictive ability for starch and dry
matter content\@.
For fat and sugar, G was the superior predictor inside the reduced set.
In the full set of 1,521 hybrids, predictive abilities obtained with G increased
markedly for all traits with the exception of fat\@.
The single-step combination of genomic and transriptomic information
($T^{*}|G$) yielded a small improvement over G alone for the traits dry matter
yield and protein.\@.
Compared to using only G, trait prediction did not benefit from the single-step
approach when T was not superior to G in the reduced set.
The only exception was protein.
Predictive abilities resulting from combinations of pedigree with genomic
($G^{*}|P$) information were higher than P alone for all traits and $T^{*}|P$
improved upon P in the `full' set for four traits.
Here, we did not consider the inclusion of dominance effects for the hybrid
material after a previous investigation into the predictive ability of various
``omics'' predictors in the same material concluded no considerable gains 
\cite{Westhues2017}.
